Introduction to\\ Swedish Civil Contingencies Agency (MSB) methodological support for introducing \ac{isms}.

### 2019-03-20 msbframework/msbintro/msbintro.tex
fixed small error
Added ToC at start of subsections.
Fixed layout and structure

\AtBeginSection[]{%
	\begin{frame}<beamer>{Overview}
		\tableofcontents[currentsection]
	\end{frame}
}

\begin{document}

@@

\section[Information Security]{Information Security}

\begin{frame}{MSB}
  \begin{itemize}
    \item Swedish Civil Contingencies Agency (MSB).
    \item \enquote{MSB is responsible for issues concerning civil protection,
        public safety, emergency management and civil defence, as long as no
        other authority has responsibility\cite[About MSB]{msbse}\@.
        }
    \item This includes protection of our information systems.
  \end{itemize}
\end{frame}

@@

\section[Methodilogical support]{MSB:s methodological support}
\subsection{Introduction to the methodilogical support}

\begin{frame}{Methodological support}
  \begin{itemize}
    \item Support for how to conduct work within information security in an
      organisation.

@@

\begin{frame}{Methodological overview}
  \begin{figure}
    \includegraphics[width=6cm]{Figures/kugghjulet_lis.png}
    \caption{Overview over the methodological support~\cite{msb_metodstod}.}
  \end{figure}
\end{frame}

\begin{frame}{Different parts of the Methodological support}
  \begin{figure}
    \includegraphics[width=\textwidth]{Figures/oversiktsbild-over-metodstegens-olika-delar-och-hur-de-relaterar-till-varandra.png}
    \caption{Different parts of the Methodological support~\cite{msb_metodstod}}
  \end{figure}
\end{frame}

\begin{frame}{ISO 27000}
  \begin{itemize}
    \item ISO 27000 is an international standard covering working with
      information security.
    \item This is not just for governmental agencies.
    \item Any organisation can certify themselves for ISO 27000
    \item MSB:s methodological support is adapted to the international
      standards.
  \end{itemize}
\end{frame}

\subsection{Identify and Analyse --- Overview}
\begin{frame}{Identify and Analyse --- Overview}
  \begin{itemize}
    \item Analyse the external contexts~\footnote{Omvärldsanalys} and the organisation.
    \item Requires knowledge about how the current situation looks
      like~\footnote{Nulägesanalys}.
      \begin{itemize}
        \item Look over existing documentation.
        \item Discuss within the organisation.
      \end{itemize}
    \item Mapping of the informational assets\footnote{informationskartläggning}.
    \item Analyse the risks connected to the informational assets.
    \item Find suitable security measurements.
    \item Existing security measurements are controlled using a GAP-analysis
  \end{itemize}
\end{frame}

\subsection{Analysing External and Legal Requirements (Omvärldsanalys)}
\begin{frame}{Identify and Analyse}
  \begin{itemize}
    \item Mapping 
      \begin{itemize}
        \item External parties needs and expectations
          \begin{itemize}
            \item Customers
            \item Suppliers
            \item Citizens
            \item Reviewers
          \end{itemize}
        \item Prerequisites
          \begin{itemize}
            \item Technical
            \item Social
            \item Environmental
            \item Political
          \end{itemize}
      \end{itemize}
  \end{itemize}
\end{frame}
\begin{frame}{Example of external parties}
  \begin{table}
    \small
    \caption{Example of external parties~\cite{msb_metodstod}}
    \begin{tabularx}{\textwidth}{X X X}
      Category & Party & Requirement, Role or Impact\\
      \toprule
      Suppliers: &
      \begin{itemize}
        \item Suppliers
        \item Sub-suppliers
        \item Outsourcing partners
        \item In sourcing partners
        \item Cloud Service Providers
        \item Business Partners
        \item Security Providers
      \end{itemize} &
      An organisation is dependent on suppliers, can for example require \newline
      how the supplier is working with information security.\\
    \end{tabularx}
  \end{table}
\end{frame}

\begin{frame}{Legal Requirements}
  \begin{description}
    \item[General Data Protection Regulation (GDPR)] Managing personal data.

@@

\begin{frame}{MSBFS 2016:1}
  5 §  Varje myndighet ska bedriva ett systematiskt och riskbaserat
  informationssäkerhetsarbete med stöd av ett ledningssystem för
  informationssäkerhet. I detta arbete ska standarderna ISO/IEC 27001:2014 och
  ISO/IEC 27002:2014 beaktas. Tillräckliga resurser ska tilldelas för
  informationssäkerhetsarbetet samt löpande och regelbunden information lämnas
  till myndighetsledningen.

  Detta innebär bland annat att en myndighet måste:
  \begin{enumerate}
    \item upprätta en informationssäkerhetspolicy och andra styrande dokument
      som behövs för myndighetens informationssäkerhet,
    \item utse en eller flera personer som leder och samordnar arbetet med
      informationssäkerhet,
    \item klassificera sin information med utgångspunkt i krav på
      konfidentialitet, riktighet och tillgänglighet,
    \item utifrån risk- och sårbarhetsanalyser och inträffade incidenter
      avgöra hur risker ska hanteras, samt besl uta om åtgärder för
      myndighetens informationssäkerhet,
    \item dokumentera granskningar och säkerhetsåtgärder av större betydelse
      som har vidtagits.
    \item
  \end{enumerate}
\end{frame}

@@

\subsection{Organisational Analysis\footnote{Verksamhetsanalys}}
\begin{frame}{Identify and Analyse --- Organisational Analysis}
  \begin{itemize}
    \item Mapping
    \begin{itemize}
      \item Internal parties needs, requirements and expectations.
      \begin{itemize}
        \item Decision makers
        \item Object owners
        \item Co-workers
        \item Support units
      \end{itemize}
      \item Prerequisites.
      \begin{itemize}
        \item Goals
        \item Strategies
        \item Organisational Culture
        \item Infrastructure
      \end{itemize}
      \item Identifies the organisations essential informational assets.
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}{Analyse Informational Assets}
  \begin{block}{Informational Assets}
    Informational assets are the organisations information and its resources
    that deals with information. For example to receive, store, process, show or
    communicate it~\cite{msb_metodstod}.
  \end{block}
\end{frame}

\begin{frame}{Dividing informational assets}
  \begin{description}
    \item[Primary] Refers to the main information, such as blue prints, logs or
      contracts.
    \item[Secondary] Refers to resources that are required to access the primary
      informational assets.
  \end{description}
  What will happen if you do not have your program that is required to read the
  closed proprietary format that the information is stored with?
\end{frame}

\begin{frame}{Organisational Analysis}
  \begin{block}{General analysis}
    Important to only perform a general analysis of the organisation.
  \end{block}
  \begin{block}{Previous Work?}
    Are there any previous analysis available?
  \end{block}
\end{frame}

@@

\subsection{Risk analysis}
\begin{frame}{Risk analysis}
  \begin{itemize}
    \item Identifies the information security risks.
    \item Can be performed over generally over the organisation, or on a
      specific object.
    \item The identified risks are then assessed based on probability and
      consequence.
  \end{itemize}
  \begin{block}{What to acheive with a risk analysis}
    \begin{itemize}
      \item What can happen?
      \item How probable is it?
      \item What is the consequences?
    \end{itemize}
    Note that the result of the risk analysis shouldn't be seen as the absolute
    truth.
  \end{block}
\end{frame}

\begin{frame}{When should the Risk Analysis be performed?}
  \begin{itemize}
    \item Not necessary to perform the risk analysis after analysing the
      organisation or external and legal requirements.
    \item If it is performed after, then the results from these can be used in
      the risk analysis.
    \item Otherwise, it is possible to rely on the existing knowledge of the
      organisation.
  \end{itemize}
\end{frame}
\begin{frame}{When should the Risk Analysis be performed?}
  Examples of when it can be suitable to perform a risk analysis.
  \begin{itemize}
    \item Procuring new services?
    \item Outsouring
    \item Organisational changes
  \end{itemize}
\end{frame}

\begin{frame}{How should the process for Risk Analysis look like?}
  \begin{block}{Already have a process?}
    If the organisation already have a process in place, then this can be used.
    Since it's already well established in the organisation.
  \end{block} 
  \begin{block}{If there isn't a process in place?}
    Some key points to think about when establishing a process for Risk
    Analysis.
    \begin{itemize}
      \item How should the process look like?
      \item What criteria for consequence and probability estimation should be
        used?
      \item What criteria for risk acceptance should be used?
      \end{itemize}
      In IS0 27001:2018 the management are responsible for making a decision
      about the process, criteria for consequence and probability levels, and
      what risk levels should be considered acceptable. 
  \end{block}
\end{frame}
\begin{frame}{Establishing a table for grading consequence}
  Create a table for help establishing consequence level connected to each risk.
  For example:
  \begin{itemize}
    \item Financial loss
    \item Negative effect
    \item Infringement
    \item Environmental
    \item Personal Safety
  \end{itemize}
\end{frame}
\begin{frame}{Establishing a table for grading consequence}
  \begin{table}
    \tiny
    \caption{Table for helping estimate risk~\cite{msb_metodstod}}
    \begin{tabularx}{\textwidth}{X | X | X | X}
    Consequence & Financial loss & Loss of reputation & Disruption in the
    organisation\\ 
    \toprule

@@

\subsection{GAP Analysis}
\begin{frame}{GAP Analysis}
  \begin{itemize}
    \item Identifies the gap between the current state of the organisation and
      what level of security the organisation wants to achieve.
    \item For example, the current state versus the actions mentioned in ISO
      27001~\cite[Bilaga A]{iso27001}
    \item Develop an action plan on how to fix the objects that aren't 
      sufficient.
  \end{itemize}
\end{frame}

\begin{frame}{GAP Analysis --- Statement of Applicability}
  Statement of Applicability (Uttalande om tillämplighet)
  \begin{itemize}
    \item What should be covered in the GAP-analysis?
    \item Select security meassures, either from previous analyses, or pick all.  
    \item Once finished, the organisation have a complete list of security
      meassures that are applicable to the organisation.
  \end{itemize}
\end{frame}
\begin{frame}{GAP Analysis --- Perform the GAP analysis}
  Based on the list of applicable security meassures
  \begin{itemize}
    \item Is the security meassure already in place?
    \item Estimate the severity of the security meassure.
    \item Describe the current status of this security meassurement.
  \end{itemize}

\subsection{The Results of the Analysis}
\begin{frame}{The result of the analysis}
  \begin{itemize}
    \item The result will act as a base when modelling the information security.
    \item Helps to assess how the organisation handles the information security.
    \item Identifying security measures.
  \end{itemize}

@@

\section{Utforma}
\begin{frame}{Classifying information}
  \begin{itemize}
    \item Method to evaluate how much an informational asset is worth
      protecting.

>>>>

\AtBeginSection[]{%
	\begin{frame}<beamer>{Overview}
		\tableofcontents[currentsection]
	\end{frame}
}
\AtBeginSubsection[]{%
	\begin{frame}<beamer>{Overview}
		\tableofcontents[currentsection, currentsubsection]
	\end{frame}
}
\AtBeginSubsubsection[]{%
	\begin{frame}<beamer>{Overview}
    \begin{center}
      \insertsubsubsectionhead
    \end{center}
	\end{frame}
}
\begin{document}

@@

\section[Information Security]{Information Security}
\subsection{Introduction to Information Security}

@@

\section[Methodilogical support]{MSB:s methodological support}
\begin{frame}{Introduction}{\insertsectionhead}
  \begin{itemize}
    \item Swedish Civil Contingencies Agency (MSB).
    \item \enquote{MSB is responsible for issues concerning civil protection,
        public safety, emergency management and civil defence, as long as no
        other authority has responsibility\cite[About MSB]{msbse}\@.
        }
    \item This includes protection of our information systems.
  \end{itemize}
\end{frame}

\begin{frame}{Introduction}{\insertsectionhead}
  \begin{itemize}
    \item Support for how to conduct work within information security in an
      organisation.

@@

\begin{frame}{Overview}{\insertsectionhead}
  \begin{figure}
    \includegraphics[width=6cm]{Figures/kugghjulet_lis.png}
    \caption{Overview over the methodological support~\cite{msb_metodstod}.}
  \end{figure}
\end{frame}

\begin{frame}{Components}{\insertsectionhead}
  \begin{figure}
    \includegraphics[width=8cm]{Figures/oversiktsbild-over-metodstegens-olika-delar-och-hur-de-relaterar-till-varandra.png}
    \caption{Different parts of the Methodological support~\cite{msb_metodstod}}
  \end{figure}
\end{frame}

\begin{frame}{ISO 27000}{\insertsectionhead}
  \begin{itemize}
    \item ISO 27000 is an international standard covering working with
      information security.
    \item This is not just for governmental agencies.
    \item Any organisation can certify themselves for ISO 27000
    \item MSB:s methodological support is adapted to the international
      standards.
  \end{itemize}
\end{frame}

\subsection{Identify and Analyse}
\begin{frame}{Overview}{\insertsubsectionhead}
  \begin{itemize}
    \item Analyse the external contexts~\footnote{Omvärldsanalys} and the organisation.
    \item Requires knowledge about how the current situation looks
      like~\footnote{Nulägesanalys}.
      \begin{itemize}
        \item Look over existing documentation.
        \item Discuss within the organisation.
      \end{itemize}
    \item Mapping of the informational assets\footnote{informationskartläggning}.
    \item Analyse the risks connected to the informational assets.
    \item Find suitable security measurements.
    \item Existing security measurements are controlled using a GAP-analysis
  \end{itemize}
\end{frame}

\subsubsection{Analysing External and Legal Requirements (Omvärldsanalys)}
\begin{frame}{Analyse External Requirements}{\insertsubsectionhead}
  \begin{itemize}
    \item Mapping 
      \begin{itemize}
        \item External parties needs and expectations
          \begin{itemize}
            \item Customers
            \item Suppliers
            \item Citizens
            \item Reviewers
          \end{itemize}
        \item Prerequisites
          \begin{itemize}
            \item Technical
            \item Social
            \item Environmental
            \item Political
          \end{itemize}
      \end{itemize}
  \end{itemize}
\end{frame}
\begin{frame}{Example of external parties}
  \begin{table}
    \footnotesize
    \caption{Example of external parties~\cite{msb_metodstod}}
    \begin{tabularx}{\textwidth}{X X X}
      Category & Party & Requirement, Role or\newline Impact\\
      \toprule
      Suppliers: &
      \begin{itemize}
        \item Suppliers
        \item Sub-suppliers
        \item Outsourcing partners
        \item In sourcing partners
        \item Cloud Service Providers
        \item Business Partners
        \item Security Providers
      \end{itemize} &
      An organisation is dependent on suppliers, can for example require \newline
      how the supplier is working with information security.\\
    \end{tabularx}
  \end{table}
\end{frame}

\begin{frame}{Legal Requirements}
  \footnotesize
  \begin{description}
    \item[General Data Protection Regulation (GDPR)] Managing personal data.

@@

\begin{frame}{MSBFS 2016:1}
  \footnotesize
  5 §  Varje myndighet ska bedriva ett systematiskt och riskbaserat
  informationssäkerhetsarbete med stöd av ett ledningssystem för
  informationssäkerhet. I detta arbete ska standarderna ISO/IEC 27001:2014 och
  ISO/IEC 27002:2014 beaktas. Tillräckliga resurser ska tilldelas för
  informationssäkerhetsarbetet samt löpande och regelbunden information lämnas
  till myndighetsledningen.

  Detta innebär bland annat att en myndighet måste:
  \begin{enumerate}
    \item upprätta en informationssäkerhetspolicy och andra styrande dokument
      som behövs för myndighetens informationssäkerhet,
    \item utse en eller flera personer som leder och samordnar arbetet med
      informationssäkerhet,
    \item klassificera sin information med utgångspunkt i krav på
      konfidentialitet, riktighet och tillgänglighet,
    \item utifrån risk- och sårbarhetsanalyser och inträffade incidenter
      avgöra hur risker ska hanteras, samt besl uta om åtgärder för
      myndighetens informationssäkerhet,
    \item dokumentera granskningar och säkerhetsåtgärder av större betydelse
      som har vidtagits.
  \end{enumerate}
\end{frame}

@@

\subsubsection{Organisational Analysis\footnote{Verksamhetsanalys}}
\begin{frame}{\insertsubsubsectionhead}
  \begin{itemize}
    \item Mapping
    \begin{itemize}
      \item Internal parties needs, requirements and expectations.
      \begin{itemize}
        \item Decision makers
        \item Object owners
        \item Co-workers
        \item Support units
      \end{itemize}
      \item Internal Prerequisites.
      \begin{itemize}
        \item Goals
        \item Strategies
        \item Organisational Culture
        \item Infrastructure
      \end{itemize}
      \item The organisations essential informational assets.
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}{Analyse Informational Assets}{\insertsubsubsectionhead}
  \begin{block}{Informational Assets}
    Informational assets are the organisations information and its resources
    that deals with information. For example to receive, store, process, show or
    communicate it~\cite{msb_metodstod}.
  \end{block}
\end{frame}

\begin{frame}{Dividing informational assets}{\insertsubsubsectionhead}
  \begin{description}
    \item[Primary] Refers to the main information, such as blue prints, logs or
      contracts.
    \item[Secondary] Refers to resources that are required to access the primary
      informational assets.
  \end{description}
  What will happen if you do not have your program that is required to read the
  closed proprietary format that the information is stored with?
\end{frame}
\begin{frame}{Critical Informational Assets}
  \centering
  \begin{block}{What is a critical informational asset?}
    There are alot of informational assets in an organisation. It is therefore
    necessary to establish what should be considered a critical asset.
  \end{block}
\end{frame}

\begin{frame}{Performing the organisational analysis}{\insertsubsubsectionhead}
  \begin{block}{General analysis}
    Important to only perform a general analysis of the organisation.
  \end{block}
  \begin{block}{Previous Work?}
    Are there any previous analysis available?
  \end{block}
\end{frame}

@@

\subsubsection{Risk analysis}
\begin{frame}{Overview}{\insertsubsubsectionhead}
  \begin{itemize}
    \item Identifies the information security risks.
    \item Can be performed over generally over the organisation, or on a
      specific object.
    \item The identified risks are then assessed based on probability and
      consequence.
  \end{itemize}
  \begin{block}{What to acheive with a risk analysis}
    \begin{itemize}
      \item What can happen?
      \item How probable is it?
      \item What is the consequences?
    \end{itemize}
    Note that the result of the risk analysis shouldn't be seen as the absolute
    truth.
  \end{block}
\end{frame}

\begin{frame}{When should the Risk Analysis be performed?}{\insertsubsubsectionhead}
  \begin{itemize}
    \item Not necessary to perform the risk analysis after analysing the
      organisation or external and legal requirements.
    \item If it is performed after, then the results from these can be used in
      the risk analysis.
    \item Otherwise, it is possible to rely on the existing knowledge of the
      organisation.
  \end{itemize}
\end{frame}
\begin{frame}{When should the Risk Analysis be performed?}{\insertsubsubsectionhead}
  Examples of when it can be suitable to perform a risk analysis.
  \begin{itemize}
    \item Procuring new services?
    \item Outsouring
    \item Organisational changes
  \end{itemize}
\end{frame}

\begin{frame}{How should the process for Risk Analysis look like?}{\insertsubsubsectionhead}
  \footnotesize
  \begin{block}{Already have a process?}
    If the organisation already have a process in place, then this can be used.
    Since it's already well established in the organisation.
  \end{block} 
  \begin{block}{If there isn't a process in place?}
    Some key points to think about when establishing a process for Risk
    Analysis.
    \begin{itemize}
      \item How should the process look like?
      \item What criteria for consequence and probability estimation should be
        used?
      \item What criteria for risk acceptance should be used?
      \end{itemize}
  \end{block}
  \begin{block}{Establish Criterias for Consequence and Probability}
    In IS0 27001:2018 the management are responsible for making a decision
    about the process, criteria for consequence and probability levels, and
    what risk levels should be considered acceptable. 
  \end{block}
\end{frame}
\begin{frame}{Establishing a table for grading consequence}{\insertsubsubsectionhead}
  Create a table for help establishing consequence level connected to each risk.
  For example:
  \begin{itemize}
    \item Financial loss
    \item Negative effect
    \item Infringement
    \item Environmental
    \item Personal Safety
  \end{itemize}
\end{frame}
\begin{frame}{Establishing a table for grading consequence}{\insertsubsubsectionhead}
  \begin{table}
    \tiny
    \caption{Table for helping estimate risk~\cite{msb_metodstod}}
    \begin{tabularx}{\textwidth}{X | X | X | X}
    Consequence & Financial loss & Loss of reputation & Disruption in the
    organisation\\ 
    \toprule

@@

\subsubsection{GAP Analysis}
\begin{frame}{Overview}{\insertsubsubsectionhead}
  \begin{itemize}
    \item Identifies the gap between the current state of the organisation and
      what level of security the organisation wants to achieve.
    \item For example, the current state versus the actions mentioned in ISO
      27001~\cite[Bilaga A]{iso27001}
    \item Develop an action plan on how to fix the objects that aren't 
      sufficient.
  \end{itemize}
\end{frame}

\begin{frame}{Statement of Applicability}{\insertsubsubsectionhead}
  Statement of Applicability (Uttalande om tillämplighet)
  \begin{itemize}
    \item What should be covered in the GAP-analysis?
    \item Select security meassures, either from previous analyses, or pick all.  
    \item Once finished, the organisation have a complete list of security
      meassures that are applicable to the organisation.
  \end{itemize}
\end{frame}
\begin{frame}{Perform the GAP analysis}{\insertsubsubsectionhead}
  Based on the list of applicable security meassures
  \begin{itemize}
    \item Is the security meassure already in place?
    \item Estimate the severity of the security meassure.
    \item Describe the current status of this security meassurement.
  \end{itemize}
\end{frame}
\subsubsection{The Results of the Analysis}
\begin{frame}{The result of the analysis}{\insertsubsubsectionhead}
  \begin{itemize}
    \item The result will act as a base when modelling the information security.
    \item Helps to assess how the organisation handles the information security.
    \item Identifying security measures.
  \end{itemize}

@@

\subsection{Utforma}
\subsubsection{Classifying Information}
\begin{frame}{Classifying information}
  \begin{itemize}
    \item Method to evaluate how much an informational asset is worth
      protecting.
